---
title: "Lab 07 — Model Selection and Validation"
author: "EAES 480 — Modern Statistics in Earth & Environmental Science"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: cosmo
    highlight: tango
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    df_print: paged
editor_options:
  chunk_output_type: inline
---

```{r setup, echo=FALSE, message=FALSE, warning=T}
library(tidyverse)
library(lubridate)
library(janitor)
library(broom)

knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)
```

# Overview

This lab extends your linear modeling work from **fitting and interpreting regression models** to **comparing models**.

In Part I, you will evaluate model fit using:

- **sum of squares**
- **R-squared**
- **MSE**
- **RMSE**
- and **residual diagnostics**

You will work with the **US-AMS simplified AmeriFlux dataset** again and choose your own response and predictor variables.

Part II will later extend this work to **train-test validation splits**.

---

# Learning goals

By the end of this lab, you should be able to:

- Fit and compare multiple linear regression models
- Use `tidy()`, `glance()`, and `augment()` to summarize models
- Compute **SST**, **SSE**, and **R-squared**
- Compute **MSE** and **RMSE**
- Compare models using both statistical summaries and plots
- Explain why model selection requires more than one metric
- Produce a fully reproducible R Markdown analysis

---

# Data

## Load and inspect

```{r load_data, echo=TRUE, eval=T}
# GOAL: Read the dataset and keep 2023 observations.

df <- read_csv("data/us-ams-simple.csv", na = c("-9999")) %>%
  clean_names() %>%
  filter(year_local == 2023)

glimpse(df)
```

## Derive time columns if needed

```{r derive_time, echo=TRUE, eval=T}
# GOAL: Create date / month columns if they do not already exist.

df <- df %>%
  mutate(
    date = as.Date(floor(doy) - 1, origin = paste0(year_local, "-01-01")),
    month = month(date, label = TRUE, abbr = TRUE)
  )
```

---

# Part I — Model selection and assessment

## 1) Choose a response variable and candidate predictors

Pick:
- **one numeric response variable**
- **two candidate predictor variables**

Possible examples:
- response: `le`, `fc`, `gpp`
- predictors: `ta`, `swc`, `vpd`, `rg`

```{r choose_vars, echo=TRUE, eval=T}
response_var <- "le"
pred1 <- "ta"
pred2 <- "swc"

# CHECK: do these variables exist?
c(response_var, pred1, pred2) %in% names(df)
```

> Which variables did you choose, and why do you think they may be related?

---

## 2) Create a clean analysis dataset

Keep only rows with non-missing values for your chosen variables.

```{r make_analysis_df, echo=TRUE, eval=T}
# GOAL: Create an analysis data frame using only complete cases.

df_mod <- df %>%
  select(response_var, pred1, pred2) %>%
  drop_na()

glimpse(df_mod)
```

---

## 3) Fit three models

Fit:
- **Model 1:** response ~ pred1
- **Model 2:** response ~ pred2
- **Model 3:** response ~ pred1 + pred2

```{r fit_models, echo=TRUE, eval=T}
m1 <- lm(df_mod[[response_var]] ~ df_mod[[pred1]], data = df_mod)
m2 <- lm(df_mod[[response_var]] ~ df_mod[[pred2]], data = df_mod)
m3 <- lm(df_mod[[response_var]] ~ df_mod[[pred1]] + df_mod[[pred2]], data = df_mod)
```

---

## 4) View coefficients with `tidy()`

```{r tidy_models, echo=TRUE, eval=T}
tidy(m1)
tidy(m2)
tidy(m3)
```

> Which coefficients are positive? Which are negative? Do they make scientific sense?

---

## 5) Visualize at least two one-predictor models

Make scatterplots with fitted linear trends for Model 1 and Model 2.

```{r plot_m1, echo=TRUE, eval=T}
ggplot(df_mod, aes(x = .data[[pred1]], y = .data[[response_var]])) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_classic(base_size = 18) +
  labs(
    x = pred1,
    y = response_var,
    title = "Model 1"
  )
```

```{r plot_m2, echo=TRUE, eval=T}
ggplot(df_mod, aes(x = .data[[pred2]], y = .data[[response_var]])) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_classic(base_size = 18) +
  labs(
    x = pred2,
    y = response_var,
    title = "Model 2"
  )
```

> Visually, which model seems stronger? Why?

---

## 6) Get fitted values and residuals with `augment()`

```{r augment_models, echo=TRUE, eval=T}
aug_m1 <- augment(m1)
aug_m2 <- augment(m2)
aug_m3 <- augment(m3)

aug_m1 %>% select(.fitted, .resid) %>% head()
```

---

## 7) Compute sum of squares manually for Model 1

Recall:

- SST = total sum of squares
- SSE = sum of squared errors
- SSR = regression sum of squares

And:

$SST = SSR + SSE$

```{r ss_model1, echo=TRUE, eval=T}
y <- df_mod[[response_var]]
ybar <- mean(y)

sst <- sum((y - ybar)^2)
sse_m1 <- sum(aug_m1$.resid^2)
ssr_m1 <- sst - sse_m1

tibble(
  sst = sst,
  sse_m1 = sse_m1,
  ssr_m1 = ssr_m1
)
```

> In words, what do SST and SSE represent?

---

## 8) Compute R-squared manually for Model 1

Use:

$R^2 = 1 - SSE / SST$

```{r r2_manual, echo=TRUE, eval=T}
r2_m1_manual <- 1 - sse_m1/sst

r2_m1_manual
```

## 9) Compare to `glance()`

```{r glance_model1, echo=TRUE, eval=T}
glance(m1)
```

> Does your manual R-squared match the value from `glance()`?

---

## 10) Compute MSE and RMSE for Model 1

Use:

- $MSE = mean(e_i^2)$
- $RMSE = sqrt(MSE)$

```{r mse_rmse_m1, echo=TRUE, eval=T}
mse_m1 <- mean(aug_m1$.resid^2)
rmse_m1 <- sqrt(mse_m1)

tibble(
  mse_m1 = mse_m1,
  rmse_m1 = rmse_m1
)
```

> What does RMSE mean in the units of your response variable?

---

## 11) Compare all three models using `glance()`

```{r compare_glance, echo=TRUE, eval=T}
bind_rows(
  glance(m1) %>% mutate(model = "m1"),
  glance(m2) %>% mutate(model = "m2"),
  glance(m3) %>% mutate(model = "m3")
) %>%
  select(model, r.squared, adj.r.squared, sigma, AIC, BIC)
```

> Which model has the highest R-squared? Is that enough to call it the best model?

---

## 12) Compare all three models using RMSE

```{r compare_rmse, echo=TRUE, eval=T}
rmse_tbl <- bind_rows(
  augment(m1) %>%
    summarise(rmse = sqrt(mean(.resid^2))) %>%
    mutate(model = "m1"),
  augment(m2) %>%
    summarise(rmse = sqrt(mean(.resid^2))) %>%
    mutate(model = "m2"),
  augment(m3) %>%
    summarise(rmse = sqrt(mean(.resid^2))) %>%
    mutate(model = "m3")
)

rmse_tbl
```

> Which model has the smallest RMSE? Does that agree with the R-squared ranking?

---

## 13) Compare residual plots

Make residual vs fitted plots for at least two models.

```{r resid_plot_m1, echo=TRUE, eval=T}
ggplot(aug_m1, aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme_classic(base_size = 18) +
  labs(
    x = "Fitted values",
    y = "Residuals",
    title = "Residuals: Model 1"
  )
```

```{r resid_plot_m3, echo=TRUE, eval=T}
ggplot(aug_m3, aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme_classic(base_size = 18) +
  labs(
    x = "Fitted values",
    y = "Residuals",
    title = "Residuals: Model 3"
  )
```

> Do the residual plots suggest any problems (curvature, changing variance, outliers)? Which model looks more appropriate?

---

## 14) Make a model selection decision

Choose the model you think is best supported by the evidence.

Fill in:

> My preferred model is `___`

> I based this decision on:
> - `___`
> - `___`
> - `___`

**Prompt (5–7 sentences):** Explain why you chose this model. Use at least two quantitative criteria (such as R-squared, RMSE, or sigma) and one visual criterion (such as the residual plot or scatterplot). Also explain why a different model was less convincing.

> *Write your answer here.*

---

# Part II — Placeholder: train-test validation

In the next lecture and lab section, we will extend model selection into **validation**.

We will ask:

- Does a model that fits the training data well also predict new data well?
- How do train-test splits change model comparison?
- When can a more complex model overfit?

> *Instructor will add this section later. Do not delete this header.*

---

# Submission

- Knit your `.Rmd` to HTML.
- Commit and push both the `.Rmd` and knitted `.html` to your GitHub repo.
- Submit the GitHub repository link on Canvas.
